---
title: "Invertebrate report: Soil macrofauna communities vary demonstrably in response to anthropogenic changes to environmental gradients in Narok County, Kenya"

author: "G.C. North, L.E. Frelich, and A.E. Guthmann"

date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

<br>
<br>

###  Data reading & cleaning

<br>

##### This step loads necessary packages, reads in all pertinent data files, and does a bit of data cleaning as well. 

<br>

```{r setup, include = TRUE, message = FALSE}

knitr::opts_chunk$set(echo = TRUE)

require(tidyverse)
require(dplyr)
require(ggplot2)
require(readr)
require(vegan)
require(car)
require(forcats)
require(ggpubr)
require(tidyr)
require(lme4)
require(cowplot)

decomp <- read_csv("DECOMP2.csv")
ants <- read_csv("ANTS2.csv")
worms <- read_csv("WORMS2.csv")
terms <- read_csv("TERMS2.csv")
soil <- read_csv("soildat.csv")


RDA_env <- read_csv("RDA_env.csv")
RDA_obs <- read_csv("RDA_obs.csv")

boma.vec <- unlist(rep(c(1039.7,
              1091.6,
              1239.9,
              4673.7,
              126.5,
              3961,
              3873.5,
              2206.9,
              2253,
              1929.3,
              2759.9,
              1983.8,
              2199.8,
              540.8,
              484.5,
              1949.5,
              410.3,
              1531.2),
              each = 5))

boma.vec <- as.data.frame(boma.vec)

scaled_nbcoefs <- read_csv("scaled_nbcoefs.csv")
scaled_logcoefs <- read_csv("scaled_logcoefs.csv")

rda.coefs <- read_csv("rda_coefs2.csv")
rda.coefs <- rda.coefs %>%
  select(,-c(6:8))

decomp <- decomp %>%
  select(-c(1)) # df containing obs for all three taxa

decomp.agg <- decomp %>%
  group_by(SiteID, PointID, hab_type, taxa) %>%
  summarise(num_ind = sum(num_ind))
    
decomp.agg <- decomp.agg %>% # Calculate relative proportions
 group_by(PointID) %>%
 mutate(proportion = num_ind / sum(num_ind))

ants <- ants %>%
  select(-c(1)) %>%
  cbind(boma.vec) # df contains ant observations only

names(ants)[names(ants) == "boma.vec"] <- "DistBoma"
ants$logDistBoma <- log(ants$DistBoma)

worms <- worms %>%
  select(-c(1)) %>%
  cbind(boma.vec) # df contains worm obs only

names(worms)[names(worms) == "boma.vec"] <- "DistBoma"
worms$logDistBoma <- log(worms$DistBoma)

terms <- terms %>%
  select(-c(1)) %>%
  cbind(boma.vec) # termite obs only

names(terms)[names(terms) == "boma.vec"] <- "DistBoma"
terms$logDistBoma <- log(terms$DistBoma)


RDA_env <- RDA_env %>%
  select(-c(1))  

RDA_env$hab_type <- as.factor(RDA_env$hab_type)

RDA_env$hab_type <- relevel(RDA_env$hab_type, 
                            ref = "grassland")

RDA_obs <- RDA_obs %>%
  select(-c(1))

rda.coefs <- rda.coefs %>%
  select(-c(1))

rda.coefs <- rda.coefs[-c(83:87), ]
```

<br>
<br>

##### Outputted objects are as follows:

- decomp (df, all observations [dim: 270 x 30])

- ants (df, ant observations only [dim: 90 x 32])

- terms (df, termite observations only [dim: 90 x 32])

- worms (df, earthworm observations only [dim: 90 x 32])

- soil (df, soil characteristics [dim: 36 x 5])

- RDA_env (df, environmental variables for RDA biplot [dim: 82 x 7])

- RDA_obs (df, observations (Hellinger-transformed abundance data for each taxon) for RDA biplot [dim: 82 x 3])

- rda.coefs (df, Euclidean distances extracted from RDA [dim: 82 x 4])

- scaled_logcoefs (df, logistic regression coefficients for all predictors and taxa, including estimate, SE, upper and lower confidence interval boundaries, and associated p-values [dim: 18 x 7])

- scaled_nbcoefs (df, negative binomial regression coefficients for all predictors and taxa, including estimate, SE, upper and lower confidence interval boundaries, and associated p-values [dim: 18 x 7])

<br>
<br>

## Figures

<br>

### Abundance by Taxa - boxplot, ANOVA, & Tukey HSD

##### Here we plot log (n+1)-transformed abundance distributions for each taxon of interest, descending by median value. Subsequently, we conducted an ANOVA and a post-hoc Tukey HSD test to assess the probability of significant differences between mean log (n+1)-transformed abundance values

```{r ANOVA and Tukey HSD}

ggplot(decomp,
       aes(x = reorder(taxa,
                       desc(logNum_ind),  # Use desc() to specify descending order
                       median, 
                       na.rm = TRUE),
           y = logNum_ind,
           fill = taxa)) +
  geom_boxplot() +
  labs(x = "Taxon",
       y = "Number individuals at site (log [n+1]-transformed)",
       fill = "Taxa") + 
  theme_minimal()

model <- aov(logNum_ind ~ taxa, 
                          data = decomp)

post_hoc <- TukeyHSD(model)

summary(model)
print(post_hoc)

```

<br>

##### Now, we break down log (n+1)-transformed abundances by habitat type

<br>

```{r, message = FALSE, warning = FALSE}

decomp2 <- decomp %>%
  group_by(hab_type, taxa) %>%
  summarise(
    sd = sd(logNum_ind, na.rm = TRUE),
    len = mean(logNum_ind, na.rm = TRUE),
    n = n(),
    se = sd/sqrt(n))

ggplot(decomp2, aes(x = taxa, 
                    y = len, 
                    fill = taxa)) +
  geom_bar(stat = "identity", 
           position = "dodge") +
  geom_errorbar(aes(ymin = len - se, 
                    ymax = len + se),
                position = position_dodge(width = 0.9),
                width = 0.25) +
  stat_summary(fun.y = "mean",
               geom = "text",
               position = position_dodge(width = 0.9),
               vjust = -0.5,
               hjust = 1.5,
               aes(label = sprintf("%.2f", ..y..)),
               size = 3,
               color = "black") +
  facet_wrap(~ hab_type,
             scales = "free",
             labeller = labeller(hab_type = c("Bush or forest",
                                              "Grassland",
                                              "Human use",
                                              "Wetland"))) +
  labs(title = "Mean site-level abundance by taxa",
       x = "Taxa",
       y = "Mean number individuals (log-transformed)",
       fill = "Taxa") +
  theme_minimal()

```

<br>
<br>

### Redundancy Analysis (RDA) & Habitat Partitioning Biplots

<br>

##### Here we conduct a constrained ordination (RDA) to visually evaluate relationships between environmental predictor variables and their Hellinger-transformed observed abundances for ants, termites and earthworms. 

<br>

``` {r, message = FALSE, warning = FALSE}

invt.rda <- rda(RDA_obs ~ ., data = RDA_env) 

summary(invt.rda)

ordiplot(invt.rda, 
         scaling = 2,  # Type 2 scaling - effects of explanatory (environmental) vars
         type = "point")

```

<br>

##### Here we replot the Euclidean values RD1 and RD2 including convex hulls around each habitat strata. 

``` {r, message = FALSE, warning = FALSE}

# stat_chull() fxn doesn't seem to want to complete the hulls; had to manually add final line segments manually

bushstart <- c(0.59195, 0.237716)
bushend <- c(0.63775, -0.147577)

grassstart <- c(0.13452, -0.36705)
grassend <- c(0.31892, 0.310784)

humanstart <- c(-0.35506, 0.866577)
humanend <- c(-0.25939, 0.404572)

wetstart <- c(0.86197, -0.304876)
wetend <- c(0.62677, -0.583837)

centroids <- data.frame(x = c(0.398124167,
                              -0.206176842,
                              -0.4457575,
                              0.420318),
                        y = c(0.084908917,
                              -0.001835684,
                              0.492854167,
                              -0.34317015),
                        hab_type = c("bush/forest",
                                     "grassland",
                                     "humanuse",
                                     "wetland"))

hab.plot <- ggplot(data = rda.coefs, aes(x = RD1,
                             y = RD2,
                             color = hab_type)) +
  geom_point() +
  geom_vline(xintercept = 0, 
             linetype = "solid", 
             color = "black") +
  geom_hline(yintercept = 0, 
             linetype = "solid", 
             color = "black") +
  coord_cartesian(xlim = c(-2, 0.75),
                  ylim = c(-1, 1)) +
  stat_chull(aes(colour = hab_type)) +  
  geom_segment(aes(x = grassstart[1],
                   y = grassstart[2],
                   xend = grassend[1],
                   yend = grassend[2]),
               color = "darkolivegreen3",
               size = 0.3,
               alpha = 0.4) +
  geom_segment(aes(x = humanstart[1],
                   y = humanstart[2],
                   xend = humanend[1],
                   yend = humanend[2]),
               color = "lightblue3",
               size = 0.3,
               alpha = 0.4) +
  geom_segment(aes(x = bushstart[1],
                   y = bushstart[2],
                   xend = bushend[1],
                   yend = bushend[2]),
               color = "coral1",
               size = 0.3,
               alpha = 0.4) +
  geom_segment(aes(x = wetstart[1],
                   y = wetstart[2],
                   xend = wetend[1],
                   yend = wetend[2]),
               color = "darkorchid1",
               size = 0.3,
               alpha = 0.4) +
  labs(color = "Habitat Type") +  
  theme_bw()

hab.plot +
  geom_point(data = centroids, 
             aes(x, y, color = hab_type), 
             size = 4,
             shape = "diamond",
             color = "black")
```


<br>
<br>

### Presence Model— Logistic Regression

<br>

#### Here we model taxon presence as a function of 6 predictor variables:

- logDistWater: Distance to Water in meters (log (n) transformed)

- logDistBoma: Distance to Boma in meters (log (n) transformed)

- logBiomass: Local Vegetation Biomass (m^3 / km^2 / year; log (n) transformed)

- PctClay: Soil Clay Content (%)

- PctWaterCt: Litter Water Content (%)

- FreshWgtCor: Litter Mass (grams)

<br>

``` {r, message = FALSE, warning = FALSE}
# converting to factor to appropriately specify SiteID as random effect
ants$SiteID <- as.factor(ants$SiteID)
terms$SiteID <- as.factor(terms$SiteID)
worms$SiteID <- as.factor(worms$SiteID)

# scale and center predictor vars
ants$logDistWater_scaled <- scale(log(ants$DistWater))
ants$PctClay_scaled <- scale(ants$PctClay)
ants$PctWaterCt_scaled <- scale(ants$PctWaterCt)
ants$logDistBoma_scaled <- scale(ants$logDistBoma)
ants$logBiomass_scaled <- scale(ants$logBiomass)
ants$FreshWgtCor_scaled <- scale(ants$FreshWgtCor)

# scale and center predictor vars
terms$logDistWater_scaled <- scale(log(terms$DistWater))
terms$PctClay_scaled <- scale(terms$PctClay)
terms$PctWaterCt_scaled <- scale(terms$PctWaterCt)
terms$logDistBoma_scaled <- scale(terms$logDistBoma)
terms$logBiomass_scaled <- scale(terms$logBiomass)
terms$FreshWgtCor_scaled <- scale(terms$FreshWgtCor)

# scale and center predictor vars
worms$logDistWater_scaled <- scale(log(worms$DistWater))
worms$PctClay_scaled <- scale(worms$PctClay)
worms$PctWaterCt_scaled <- scale(worms$PctWaterCt)
worms$logDistBoma_scaled <- scale(worms$logDistBoma)
worms$logBiomass_scaled <- scale(worms$logBiomass)
worms$FreshWgtCor_scaled <- scale(worms$FreshWgtCor)

log.ant.model <- glmer(presence ~ logDistWater_scaled + 
                         logBiomass_scaled + 
                         logDistBoma_scaled + 
                         PctWaterCt_scaled + 
                         PctClay_scaled + 
                         FreshWgtCor_scaled + 
                         (1 | SiteID),
                       data = ants, 
                       family = binomial, 
                       control = glmerControl(optimizer = "bobyqa"),
           nAGQ = 10)

log.term.model <- glmer(presence ~ logDistWater_scaled + 
                         logBiomass_scaled + 
                         logDistBoma_scaled + 
                         PctWaterCt_scaled + 
                         PctClay_scaled + 
                         FreshWgtCor_scaled + 
                         (1 | SiteID),
                       data = terms, 
                       family = binomial, 
                       control = glmerControl(optimizer = "bobyqa"),
                       nAGQ = 10)

log.worm.model <- glmer(presence ~ logDistWater_scaled + 
                         logBiomass_scaled + 
                         logDistBoma_scaled + 
                         PctWaterCt_scaled + 
                         PctClay_scaled + 
                         FreshWgtCor_scaled + 
                         (1 | SiteID),
                       data = ants, 
                       family = binomial, 
                       control = glmerControl(optimizer = "bobyqa"),
                       nAGQ = 10)

```

<br>
<br>

##### And we output three logistic regression models:

<br>

``` {r, message = FALSE, warning = FALSE}

summary(log.ant.model)

summary(log.term.model)

summary(log.worm.model)

```

<br>
<br>

### Abundance Model— Negative Binomial Regression

<br>

##### Here we model taxon abundance as a function of the *same* 6 predictor variables:

- logDistWater: Distance to Water in meters (log (n) transformed)

- logDistBoma: Distance to Boma in meters (log (n) transformed)

- logBiomass: Local Vegetation Biomass (m^3 / km^2 / year; log (n) transformed)

- PctClay: Soil Clay Content (%)

- PctWaterCt: Litter Water Content (%)

- FreshWgtCor: Litter Mass (grams)

<br>

##### Note: A negative binomial regression model was selected given our overdispersed count data. 

<br>

``` {r, message = FALSE, warning = FALSE}

logants.nb.model <- glmer.nb(logNum_ind ~ logDistWater_scaled + 
                                logBiomass_scaled + 
                                logDistBoma_scaled + 
                                PctWaterCt_scaled + 
                                PctClay_scaled + 
                                FreshWgtCor_scaled + 
                                (1 | SiteID), 
                              data = ants,
                              family = negbinomial(link = "log"),
                              control = glmerControl(optimizer = "bobyqa"))

logterms.nb.model <- glmer.nb(logNum_ind ~ logDistWater_scaled + 
                                logBiomass_scaled + 
                                logDistBoma_scaled + 
                                PctWaterCt_scaled + 
                                PctClay_scaled + 
                                FreshWgtCor_scaled + 
                                (1 | SiteID), 
                              data = terms,
                              family = negbinomial(link = "log"),
                              control = glmerControl(optimizer = "bobyqa"))


logworms.nb.model <- glmer.nb(logNum_ind ~ logDistWater_scaled + 
                                logBiomass_scaled + 
                                logDistBoma_scaled + 
                                PctWaterCt_scaled + 
                                PctClay_scaled + 
                                FreshWgtCor_scaled + 
                                (1 | SiteID), 
                              data = worms,
                              family = negbinomial(link = "log"),
                              control = glmerControl(optimizer = "bobyqa"))

```

<br>
<br>
<br>

##### And we output three negative binomial regression models:

``` {r, message = FALSE, warning = FALSE}

summary(logants.nb.model)
summary(logterms.nb.model)
summary(logworms.nb.model)

```

<br>
<br>

### Forest Plots: Logistic & Negative Binomial Regression Coefficients

<br>

##### Using the above models, we extracted the logistic and negative binomial regression model coefficients (objects scaled_logcoefs and scaled_nbcoefs from data import)

<br>

``` {r, message = FALSE, warning = FALSE}

# Logistic regression model forest plot 
scaled_logcoefs %>% 
  ggplot(aes(x = variable,
             y = estimate,
             color = taxa)) +
  geom_point(size = 2.5,
             position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = lower_ci, 
                    ymax = upper_ci, 
                    color = taxa), 
                width = .25,
                size = 1,
                position = position_dodge(width = 0.5)) +
  scale_color_manual(name = "Taxon", 
                     labels = c("Ants", 
                                "Termites", 
                                "Earthworms"),
                     values = c("coral1", 
                                "darkturquoise", 
                                "darkolivegreen3")) +
  geom_hline(yintercept = 0) +
  theme_bw() +
  theme(axis.text.x = element_text(size = 10,
                                   #angle = 30,
                                   #vjust = .7,
                                   color = "black"),
        axis.title.y = element_text(size = 12),
        axis.text.y = element_text(size = 10,
                                   color = "black"),
        legend.position = "none") +
  facet_wrap(~taxa) +
  ylab("Predictor variable coefficient estimates on taxa presence") +
  xlab("Predictor Variable") + 
  coord_flip()

```

<br>
<br>

``` {r, message = FALSE, warning = FALSE}

# Neg bin regression model forest plot 
scaled_nbcoefs %>% 
  ggplot(aes(x = variable,
             y = estimate,
             color = taxa)) +
  geom_point(size = 2.5,
             position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = lower_ci, 
                    ymax = upper_ci, 
                    color = taxa), 
                width = .25,
                size = 1,
                position = position_dodge(width = 0.5)) +
  scale_color_manual(name = "Taxon", 
                     labels = c("Ants", 
                                "Termites", 
                                "Earthworms"),
                     values = c("coral1", 
                                "darkturquoise", 
                                "darkolivegreen3")) +
  geom_hline(yintercept = 0) +
  theme_bw() +
  theme(axis.text.x = element_text(size = 10,
                                   #angle = 30,
                                   #vjust = .7,
                                   color = "black"),
        axis.title.y = element_text(size = 12),
        axis.text.y = element_text(size = 10,
                                   color = "black"),
        legend.position = "none") +
  facet_wrap(~taxa) +
  ylab("Predictor variable coefficient estimates on taxa abundance") +
  xlab("Predictor Variable") + 
  coord_flip()

```

<br>
<br>
<br>

### Community dynamics— Shannon diversity, species richness, & evenness

<br>

##### Here we investigate the relationship between soil macrofauna (i.e., ants, termites, and earthworms) abundance (log [n+1]-transformed) and various point-level soil macrofauna community metrics (i.e., Shannon diversity, species richness, and evenness). Since we have no evidence to assume that these relationships are linear, we applied a locally-weighted (loess) method for fitting our regression line. Parameter 'span' was set to 1.0 to reduce the potential for overfitting. 

<br>

``` {r, message = FALSE, warning = FALSE}

# Shannon Diversity, locally-weighted (loess) scatterplot

ggplot(data = decomp, aes(x = logNum_ind, 
                          y = ShanDiv)) +
  geom_jitter(aes(color = taxa,
                  size = 1.5)) +
  facet_wrap(~taxa) +
  xlab("Point-level Abundance (log (n+1)-transformed)") +
  ylab("Shannon Diversity") +
  geom_smooth(method = "loess", 
              span = 1,
              color = "black") +
  theme_minimal()
```

<br>

``` {r, message = FALSE, warning = FALSE}

# Species richness, locally-weighted (loess) scatterplot

ggplot(data = decomp, aes(x = logNum_ind, 
                          y = SpecRich)) +
  geom_jitter(aes(color = taxa,
                 size = 1.5)) +
  facet_wrap(~taxa) +
  xlab("Point-level Abundance (log (n+1)-transformed)") +
  ylab("Species Richness") +
  geom_smooth(method = "loess", 
              span = 1,
              color = "black") +
  theme_minimal()
```

<br>

``` {r, message = FALSE, warning = FALSE}

# Evenness, locally-weighted (loess) scatterplot

decomp <- decomp %>%
mutate(Evenness = ShanDiv / log(SpecRich))

ggplot(data = decomp, aes(x = logNum_ind, 
                          y = Evenness)) +
  geom_jitter(aes(color = taxa,
                  size = 1.5)) +
  facet_wrap(~taxa) +
  xlab("Point-level Abundance (log (n+1)-transformed)") +
  ylab("Evenness") +
  geom_smooth(method = "loess", 
              span = 1,
              color = "black") +
  theme_minimal()

```

<br>
<br>
<br>

## Supplemental Figures

<br>

##### Below are code chunks and visualization outputs associated with the supplemental figures included in our text.

<br>

### Intra-site variability in soil clay content & litter mass

<br>

##### Here we generate boxplots which characterize the intra-site variability present between points for each sample site with respect to proportion soil clay content and point litter mass (g).

<br>

``` {r, message = FALSE, warning = FALSE}

ggplot(soil, aes(x = SiteID, 
                 y = avg,
                 fill = variable)) +
  geom_bar(stat = "identity",
           position = "dodge") +
  geom_errorbar(aes(ymin = avg - sterr, 
                    ymax = avg + sterr),
                width = 0.25) +
  labs(x = "Proportion Soil Clay Content",
       y = "Mean Value") +
  scale_fill_manual(values = c("clay" = "gray", 
                               "litter" = "gray60")) +
  facet_wrap(~ variable,
             scales = "free",
             labeller = labeller(hab_type = c("Soil Clay Content (%)",
                                              "Point Litter Mass (g)"))) +
  theme_minimal()

```

<br>
<br>

### Modelling relationship between distance to water and local vegetation biomass

<br>

##### Here we investigate the relationship between distance to water (log, meters) and local vegetation biomass (m^3 / km^2 / year). Since we have no evidence to assume that these relationships are linear, we applied a locally-weighted (loess) method for fitting our regression line. Parameter 'span' was set to 1.0 to reduce the potential for overfitting. 

<br>

``` {r, message = FALSE, warning = FALSE}

ggplot(data = decomp, aes(x = DistWater, y = logBiomass)) +
  geom_point(aes(color = hab_type,
                 size = 1.5)) +
  xlab("Distance to Water (m)") +
  ylab("Local Vegetation Biomass (m3 / km2 / year)") +
  geom_smooth(method = "loess", 
              span = 1.0,
              color = "black") +
  theme_minimal()

```

<br>
<br>

### Proportions of ants, termites & earthworms by site

<br>

##### Here we model point-level community proportions for ants, termites, and earthworms using pie charts. Each pie represents one sample point; each row of five pies is representative of one sample site. Each point is labeled by habitat type. 

<br>

``` {r, message = FALSE, warning = FALSE}
 
ggplot_list <- lapply(unique(decomp.agg$PointID), function(point) {
  ggplot(decomp.agg[decomp.agg$PointID == point, ], 
         aes(x = "", y = proportion, fill = taxa)) +
    geom_bar(stat = "identity", width = 1, color = "white") +
    coord_polar("y") +
    theme_void() +
    facet_grid(~ hab_type) +
    theme(legend.position = "none") +
    scale_fill_discrete(name = "Taxa") # +
    # ggtitle(paste(decomp.agg$hab_type))
})

plot_grid(plotlist = ggplot_list, ncol = 10)

```

<br>
<br>
<br>
<br>
<br>
<br>


#### *Version Information:*
```{r, echo = FALSE}
sessionInfo()
```